\chapter*{Origin Story}
This text arose from an attempt at a formalization of Robin Hartshorne's \emph{Foundations of Projective Geometry}
using the Isabelle proof assistant, primarily relying on Isar. That project got started in 2018, continued through a small course in 2020 (interrupted by the pandemic), and restarted in 2023. As a former mathematician, I chose Hartshorne's book because of its relative simplicity -- a projective geometry is a far simpler object than, say, a field or a module, and at least in its simplest form doesn't require any algebraic objects at all -- just four simple axioms about a pair of sets. It seemed as if it should be easy to formalize, especially for someone who happened to already know some ML (the language underlying Isabelle) and who was once a mathematician, hence familiar with proofs. 

It wasn't easy after all, but the problem wasn't either the complexity of projective geometries or my experience with ML and mathematics -- it was a lack of a text that told me enough about Isabelle to let me learn it systematically. As I've gradually begun to do so (with a great deal of help, and some other sets of notes like this one), I've begun writing this book, making sure to retain my innocence: a proper degree of confusion about what to do next, something that experienced Isabelle users lack. They say things like ``this is an ordinary co-induction!" or ``prove that `there-exists' proposition by just providing a witness," but they fail to realize how very much is wrapped up in such a telegraphic suggestion. Hence I've written this rather wordy and slow-moving introduction. 
